
---
title: '**MA415 Midterm Project: Safety in the workplace in Massachusetts**'
author: "Sarah Gore- U72145380"
date: 'This version: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  prettydoc::html_pretty:
    highlight: vignette
    theme: architect
    toc: no
  pdf_document:
    highlight: tango
    number_sections: yes
    toc: yes
  html_notebook:
    code_folding: hide
    number_sections: no
    theme: flatly
    toc: yes
    toc_float: yes
subtitle: '*Making data ready for analysis*'
---
Available versions of this document: [html--notebook version](http://htmlpreview.github.com/?https://github.com/sezzagore/Midterm/blob/master/Midterm_Project.nb.html),  [pdf version](https://github.com/sezzagore/Midterm/blob/master/Midterm_Project.pdf), [R script alone](https://github.com/sezzagore/Midterm/blob/master/Midterm_Project.R), [html with full code version](http://htmlpreview.github.com/?https://github.com/sezzagore/Midterm/blob/master/Midterm_Project.html). These files can be found on my [GitHub](https://github.com/sezzagore/Midterm). 

```{r code=readLines(knitr::purl('Midterm_Project.Rmd', documentation = 1)), eval = FALSE, include=FALSE}
# the code in the description of the chunck automatically creates/saves a R script from this .Rmd
# for some reason, this seems to disallow naming further chuncks
```
---

# General purpose

## Driving Question and Present Contribution 

What is the safest workplace in MA and what can the data provided tell us about it?

I am cleaning the data provided in order to get it ready for analysis. Through doing this, I exclude data that I think is unnecessary,but at the same time I justify (to a certain extent) how and why this is done. This work is the important step between the data collection and data analysis. 

This project will only focus on the second of the following steps in a data project:

1. Data collection
2. **Data cleaning**
3. Analysis 

I acknowledge that there are many different ways to perform this project. I believe however, that this will prove useful. 

# Data

## Source
The data was collected by [OSHA](https://www.osha.gov/oshstats/index.html)  and provided by [NICAR](http://www.ire.org/nicar/). 

The data provided covers the period 01/1972- 02/2006. It contains various tables ranging from the time that the inspections were carried out, accidents incurred, body parts injured, violations, fines etc... 

The main data table is OSHA and the various supplementary tables provide key information for specific issues with the inspection such as accidents or violations, for example. There are also look-up tables that provide further sources of information about the coding for instance. 

The links between the various data files can be made more clear by consulting the following [file](https://ire.org/media/uploads/files/datalibrary/samplefiles/OSHA%20Workplace%20Safety%20Data/osha.pdf).

For more information about the variables used in the data, click on the following 
[file](http://ire.org/media/uploads/files/datalibrary/samplefiles/OSHA%20Workplace%20Safety%20Data/OSHA.txt). 

# Tables under consideration 

I will be using OSHA as it is our main source. Furthermore, I will be using ACCID as well as VIOL. In this project I am concerned with safety in the workplace and it seems reasonable to analyze data related to accidents and violations. 

I have excluded potentially useful data, such as company debt. However given the abundance of data, I find it is more relevance to choose the variables that I have. 

# Reading the data 

```{r , message=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, cache = TRUE)
require(ggplot2) 
require(tidyr)
require(dplyr)
require(readr) # to read text
require(foreign) # library used to import the dbfs
require(lubridate) # library used for dates/times
```

First I load the relevant libraries that I use in the cleaning process. 
As mentioned above, I am working with three files, `osha`, `accid` and `viol`. In order to simplify the cleaning process and upcoming analysis, I am going to merge them into one file. 

```{r, warning=FALSE}
# read the data and transform in tibble (for speed and better behaviour of the data frame)
osha <-  read.dbf("osha.dbf", as.is = TRUE) %>%
  tbl_df 
accid <- read.dbf("accid.dbf", as.is = TRUE) %>%
  tbl_df
viol <- read.dbf("viol.dbf", as.is = TRUE) %>%
  tbl_df

join_1 <- left_join(osha, accid, by = "ACTIVITYNO")
osha <- left_join(join_1, viol, by = "ACTIVITYNO")
rm(list = setdiff(ls(), "osha")) # I erase my environment and now only have osha in my environment 
```

I merge the three variables into one larger data frame and call it `osha`. This file contains all the information that was in `osha`, `accid` and `viol`. 

However, after careful consideration, I realized that this goes against the principles of tidy data. So, I keep the three data frames separated. 

```{r}
osha <-  read.dbf("osha.dbf", as.is = TRUE) %>%
  tbl_df 
accid <- read.dbf("accid.dbf", as.is = TRUE) %>%
  tbl_df
viol <- read.dbf("viol.dbf", as.is = TRUE) %>%
  tbl_df
```

## Some notes on the data

The following is an expendable collection of notes that help the reader understand the data.

  - Observations in `accid` are values for a single person in an accident. below, when counting accidents, I am referring to the number of people involved in accidents. 
  - After listening to the presentations of fellow students, I wanted to double check that the activity number's for each observation were unique before submitting my project and as I initally suspected, they were. There are `r nrow(osha)` rows in `osha` and `r length(unique(osha$ACTIVITYNO))` different inspection identifiers in the data set. They equal up to a tiny difference. Therefore NO duplicates. 
  - All date variables are in the format of `package::lubridate`. 
  
# Removing variables

At this stage I want to remove some unnecessary variables. I explain below which variables I remove and the reason for doing so. 

## Single value variables

Single value variables are vectors that contain only one unique value or one value and `NA's. In order to simplify the data sets, I remove these variables from the data frames.

For that purpose I wrote a specific function: `clean.from.one.value()`. This function takes a data frame as input and returns the same data frame without these variables.

```{r}
clean.from.one.value <- function(df){
  if(!is.data.frame(df)){
    print("Not executed: the function requires a data frame object.")
    return()
  }
  one.value <- function(vec){length(unique(vec))==1}
  clean.df <- df[,!sapply(df, one.value)]
  if(length(df)-length(clean.df)==0) {
    print("No variable removed.")
  } else {print(paste(length(df)-length(clean.df),"variables removed:",  length(df),"->",length(clean.df) ))
    }
  return(clean.df)
}
```

I now apply the function defined above to my three data frames. 

```{r, results= "hold"}
osha <- clean.from.one.value(osha)
accid <- clean.from.one.value(accid)
viol <- clean.from.one.value(viol)
```


## Irrelevant variables in `osha` 

Here I make a choice of variables that I think are unnecessary and can therefore be removed. 

### NICAR dates
NICAR converted the original dates into a standard format. Therefore, I discard the original variables that contain the original dates. 

```{r}
irrelavant_variables <- c("OSHA1MOD", "OPENDATE", "CLOSEDATE", "CLOSEDATE2",  "FRSTDENYN", "LSTREENTRN",
 "PENDUDATE", "FTADUDATE", "FRSTCONTST")

osha <- osha[,!(names(osha) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

### Number of records
`osha` contains variables indicating counts in tables that I have discarded for this project. Therefore, I discard the variables relating to these tables. 

```{r}
irrelavant_variables <- c("PROG_", "RELACT_","OPTINFO_", "DEBT_", "EVENT_", "HAZSUB_", "ADMPAY_")

osha <- osha[,!(names(osha) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

### Paperwork times  
Time spent on preparation, traveling, researching etc... are variables that I think are not going to be relevant in the analysis. 

```{r}
irrelavant_variables <- c("PAPREP" ,"PATRAVEL" ,"PAONSITE", "PATECHSUPP" ,"PARPTPREP","PAOTHRCNF","PALITIGTN","PADENIAL" ,"PASUMHOURS")

osha <- osha[,!(names(osha) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

### Debt-related variables 
Since I have discarded the tables relating to debt, I can remove these variables from `osha`. 

```{r}
irrelavant_variables <- c("PENDUDATE", "PENDUDT", "FTADUDATE", "FTADUDT", "DUECODE")

osha <- osha[,!(names(osha) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

### Penalty- related variables 
The same argument as mentioned in the previous point applies to this category. 
```{r}
irrelavant_variables <- c("PENREMIT", "FTAREMIT", "TOTPENLTY", "TOTALFTA")

osha <- osha[,!(names(osha) %in% irrelavant_variables)]
rm(irrelavant_variables)
```


### Other irrelevant variables 
This category applies to administrative information that I cannot use. 
```{r}
irrelavant_variables <- c("DUNSNO", "HOSTESTKEY")

osha <- osha[,!(names(osha) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

## Irrelevant variables in `accid`

### Unused categories  
This subset applies to variables in `accid` that I discard. 
```{r}
irrelavant_variables <- c("SITESTATE", "NAME", "RELINSP", "SOURCE", "ENVIRON", "EVENT", "TASK", "HAZSUB", "OCC_CODE")

accid <- accid[,!(names(accid) %in% irrelavant_variables)]
rm(irrelavant_variables)
```


## Irrelevant variables in `viol`

Again, I only use a subset of the variables in `viol` and discard the others. Here, the justification is that I don't think that these variables are related to safety in the workplace, for instance abatement. 

### Dates, citations and abatements etc... 

```{r}
irrelavant_variables <- c("SITESTATE", "DELETE", "ISSUANCE", "ISSUEDATE", "CITATION", "ITEMNO", "ITEMGROUP", "EMPHASIS", "PENCURRENT", "PENINITIAL", "STD_LOOKUP", "STD", "ABATE", "DATE_ABATE", "ABATEDT", "ABATEDT2", "REC", "ABATEDONE")

viol <- viol[,!(names(viol) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

### Violated Contested Information

```{r}
irrelavant_variables <- c("ERCONTDT", "ERCONDATE", "VIOLCONT", "PENCONT", "EMPRCONT", "EMPECONT", "FINORDT", "FINORDATE", "PMA", "AMENDED", "ISA", "DISPEVT" )

viol <- viol[,!(names(viol) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

### Failure to abate 

```{r}
irrelavant_variables <- c("FTAINSP", "FTAPEN", "ISSUDT", "FTA_ISDT", "CONTDT", "CONTDATE", "FTA_AMN", "FTA_ISA", "FTA_DISP", "FTA_FIN", "FTAFINDT", "HAZCAT")

viol <- viol[,!(names(viol) %in% irrelavant_variables)]
rm(irrelavant_variables)
```

# Adding information

In `osha` many variables come with coded-values; this is not very informative in particular when cleaning and graphing the data. In order to make the information more 
informative, I use the look-ups provided to add to translate the code to values I can use. 

For instance, if I want to know about the places of inspections, I will use the name of the counties and I access these through the look-ups. 


## New variables to `osha` 

I added information about the place of inspection to `osha`. *(Notice that this method is drastically different from what was shown in class)*

```{r, warning = FALSE}
# I load the lookup from scc 

scc <-  read.dbf("lookups/scc.dbf") %>%
  tbl_df %>%
  filter(STATE=="MA")
counties <-  read.dbf("lookups/scc.dbf") %>%
  tbl_df %>% filter(STATE=="MA") %>%
  filter(CITY == "0000", COUNTY!="000") %>%
  select(COUNTY, NAME) %>%
  rename(COUNTY_CODE =COUNTY, COUNTY_NAME= NAME) 
osha <- left_join(osha,counties, by=c("SITECNTY"="COUNTY_CODE"))
cities <- read.dbf("lookups/scc.dbf") %>%
  tbl_df %>%
  filter(STATE=="MA") %>%
  filter(CITY != "0000", COUNTY!="000") %>%
  select(CITY, NAME) %>%
  rename(CITY_CODE =CITY, CITY_NAME= NAME)
osha <- left_join(osha,cities, by=c("SITECITY"="CITY_CODE"))
rm(scc, counties, cities)
```

##  New variables to `accid`

I added information about the body- parts, nature of injury and human contribution in the accidents to `accid`.

```{r, warning = FALSE}
# I load the lookup from acc 

acc <-  read.dbf("lookups/acc.dbf") %>%
  tbl_df %>%
  filter(CATEGORY=="PART-BODY") %>%
  rename(BODYPART_NAME = VALUE)
accid <- left_join(accid,acc, by=c("BODYPART"="CODE"))

acc <-  read.dbf("lookups/acc.dbf") %>%
  tbl_df %>%
  filter(CATEGORY=="NATURE-INJ") %>%
  rename(NATURE_NAME = VALUE)
accid <- left_join(accid,acc, by=c("NATURE"="CODE"))

acc <-  read.dbf("lookups/acc.dbf") %>%
  tbl_df %>%
  filter(CATEGORY=="HUMAN-FAC") %>%
  rename(HUMAN_NAME = VALUE)
accid <- left_join(accid,acc, by=c("HUMAN"="CODE"))
rm(acc)
```

# NA's and missing values 

Many variables contain NA's. However there are various reasons for these NA's. In some of the cases they do represent missing values. However, when going through the descriptions of the `osha` data, it became clear that some NA's actually represent data and I want to account for that data. 

## NA's that represent actual data 

Consider the following example taken from the description of the coding:

> WALKAROUND 	C	1		Employee representative present during inspection:
B (blank) = no
X = yes

Clearly, there is no information missing. The NA seen in the data represents the value "no". Therefore in the variables that use this coding system, I replace the NA's by "no". My choice must be taken into account for future analyses.

```{r}
blank_variables <- c("CONTFLAG", "STFLAG", "PREVCTTYP", "WALKAROUND", "INTRVIEWD", "CLOSECASE", "SAFETYMANF", "SFTYCONST", "STFYMARIT", "HELTHMANF", "HELTHCONST", "HELTHMARIT", "MIGRANT", "ANTCSRVD")
blank_variables <- blank_variables[blank_variables %in% colnames(osha)]
osha[,blank_variables][is.na(osha[blank_variables])] <- "NO"
rm(blank_variables)
```

## Values that are actually NA's 

As a good practice, I plot the some of the data to look for outliers. Within the `accid` data under the age category, I encounter a bunch of zero's which does not make sense; indeed as such a large number of zeros is an indication that something is wrong. 

```{r}
accid %>%
  ggplot(aes(x = AGE)) +
  geom_bar(alpha = 0.7) +
  labs(title = "Distribution of Age of Workers in Accidents Prior to Correction", x = "Age", y = "Number of Workers")
```

I correct for this by specifically assigning the value NA to these zeros. 

```{r}
accid$AGE[accid$AGE == 0] <- NA
```

In the amended distribution, there are still many extreme values, ranging from `r min(accid$AGE)` to `r max(accid$AGE)`. So can assume that these are outliers. For instance young/old people present when the accident occurred. 

```{r, warning=FALSE}
accid %>%
  ggplot(aes(x = AGE)) +
  geom_bar(alpha = 0.7) + 
  labs(title = "Distribution of Age of Workers in Accidents After Correction", x = "Age", y = "Number of workers")
```


# First glimpse at the data 

There are very many ways of looking at the data that I have cleaned. Below are a few graphs that I feel give one a notion of the data.  

## The number of accidents per county 
Since, I am looking for secure work environment, a natural perspective would be to break-up accidents by county. 
```{r}
osha %>%
  group_by(COUNTY_NAME) %>%
  summarise(accidents = sum(ACCID_)) %>%  
  filter(!is.na(COUNTY_NAME)) %>%
  arrange(-accidents) %>%
  mutate(COUNTY_NAME = factor(COUNTY_NAME, COUNTY_NAME)) %>%
  ggplot(aes( x = COUNTY_NAME, y = accidents)) +
  geom_bar(alpha = 0.7, stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Number of Accidents Per County (1972-2006)", x = "County", y = "Number of Accidents")
```

## The number of accidents per city
I can also break-up accidents by city. However, there are too many so I take the top 15 cities in MA.
```{r}
osha %>%
  group_by(CITY_NAME) %>%
  summarise(accidents = sum(ACCID_)) %>%  
  filter(!is.na(CITY_NAME)) %>%
  arrange(-accidents) %>%
  top_n(n = 15, accidents) %>%
  mutate(CITY_NAME = factor(CITY_NAME, CITY_NAME)) %>%
  ggplot(aes( x = CITY_NAME, y = accidents)) +
  geom_bar(alpha = 0.7, stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Number of Accidents in the Top 15 Cities (1972-2006)", x = "City", y = "Number of Accidents")
```

## Accidents over time 
I am interested in the evolution in the number of accidents over time. Here I plot the total number of accidents during that period in MA. 

Notice that I exclude 1972 and 2006 because they are incomplete years. 

From the data I see that the max number of accidents in this time period was `r max(osha$ACCID_)` and the min number of accidents in this time period was `r min(osha$ACCID_)`
```{r}
osha %>%
  group_by(year(OPENDT)) %>%
  summarise(accidents = sum(ACCID_), year = mean(year(OPENDT))) %>%
  filter(year >1972, year < 2006) %>%
  ggplot(aes(x = year, y = accidents)) +
  geom_line() + labs(title = "Number of Accidents over Time (1973-2005)", x = "Year", y = "Number of Accidents")
```

I get further inside by breaking up the evolution over time by county. 

```{r}
osha %>%
  group_by(year(OPENDT), COUNTY_NAME) %>%
  filter(!is.na(COUNTY_NAME)) %>%
  summarise(accidents = sum(ACCID_), year = mean(year(OPENDT))) %>%
  filter(year >1972, year < 2006) %>%
  ggplot(aes(x = year, y = accidents, col = COUNTY_NAME, group = COUNTY_NAME)) +
  geom_line() +
  labs(title = "Number of Accidents over Time (1973-2005)", x = "Year", y = "Number of Accidents", col = "County" )
```
## Violations over time 
Violations tell us how well firms are complying with the rules of safety. This is why evolution over time is informative. Here I plot the total number of violations during that period in MA. 

Again I exclude 1972 and 2006 because they are incomplete years. 

From the data I see that the max number of accidents in this time period was `r max(osha$VIOLS_)` and the min number of accidents in this time period was `r min(osha$VIOLS_)`
```{r}

osha %>% 
  group_by(year(OPENDT)) %>%
  summarise(violations = sum(VIOLS_), year = mean(year(OPENDT))) %>%
  filter(year >1972, year < 2006) %>%
  ggplot(aes(x = year, y = violations)) +
  geom_line() +
  labs(title = "Number of Violations over Time (1973-2005)", x = "Year", y = "Number of Violations")
```

## Age of the victims of accidents 
Age usually plays a role in many work-place accidents. here I plot the distribution of age, differentiated by gender. 

```{r}

accid %>%
  filter(!is.na(SEX)) %>%
  ggplot(aes(x = SEX, y= AGE, fill = SEX)) +
  geom_boxplot() +
  labs(title = "Distribution of Age of Victims by Gender", x = "Sex", y = "Age") + guides(fill = FALSE)
```

## Nature of the injuries 

The nature of the injury can provide useful informative in terms of possible prevention.

I chose the top 15 body-parts as there are `r length(unique(accid$BODYPART))` different categories of body-parts that were harmed in the accidents.  

```{r}
accid %>%
  group_by(BODYPART_NAME) %>%
  summarise(cases = n())  %>%
  arrange(-cases)%>% top_n(n = 15, cases) %>%
  mutate(BODYPART_NAME = factor(BODYPART_NAME, BODYPART_NAME)) %>%
  filter(!is.na(BODYPART_NAME))  %>%
  ggplot(aes( x = BODYPART_NAME, y = cases)) +
  geom_bar(alpha = 0.7, stat = "identity") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Breakdown of Top 15 Body-Parts Harmed", x = "Body-Part Harmed in the Accident", y = "Number of Cases")
```


* I acknowledge that I benefitted from the help of a tutor with whom I work regularly. He reviewed my code and offered some suggestions of improvement. I remain soley responsible of this work. 

* I hope you found this as interesting as I did!


# Appendix: Variable  Description 

The following tables provide information about the variables that remain in the data set after cleaning. For further information, please refer to the original document such as [osha.txt](http://ire.org/media/uploads/files/datalibrary/samplefiles/OSHA%20Workplace%20Safety%20Data/OSHA.txt). 

## Description of variables in `osha`

```{r}
text <- read_lines("layouts/OSHA.txt", skip = 10)
text1 <- gsub("\\t\t\t.*","", text )
text2 <- text1[grepl("\t", text1)]
text3 <- gsub("\\t\t","\t", text2) %>%
  strtrim(100) %>%
  tbl_df %>%
  separate(value, c("variable", "type", "length", "description"), sep = "\t" )
text3$variable <- gsub(" ", "", text3$variable, fixed = TRUE) 
text3 <- text3  %>%
  select(variable, description) %>%
  arrange(variable) 
osha_descriptions <- text3 %>%
  filter( variable %in% names(osha))
knitr::kable(osha_descriptions, caption = "Description of Variables in `osha`")
```

## Description of variables in `accid`

```{r}
text <- read_lines("layouts/ACCID.txt", skip = 9)
text1 <- gsub("\\t\t\t\t.*","", text )
text2 <- gsub("\\t\t","\t", text1 )
text2 <- gsub("\\t\t","\t", text2 )
text2 <- gsub("\t$","", text2 )
text3 <- text2[grepl("\t", text2)]
text3 <-  text3 %>% tbl_df %>% separate(value, c("variable", "type", "length", "description"), sep = "\t" ) %>% select(variable, description) %>% arrange(variable) 
accid_descriptions <- text3 %>% filter( variable %in% names(accid)) 
knitr::kable(accid_descriptions, caption = "Description of Variables in `accid`")
```

## Description of variables in `viol` 

```{r}
text <- read_lines("layouts/VIOL.txt", skip = 7)
text1 <- gsub("\\t\t\t\t.*","", text )
text2 <- gsub("\\t\t","\t", text1 )
text2 <- gsub("\\t\t","\t", text2 )
text2 <- gsub("\t$","", text2 )
text3 <- text2[grepl("\t", text2)]
text3 <-  text3 %>% tbl_df %>% separate(value, c("variable", "type", "length", "description"), sep = "\t" ) %>% select(variable, description) %>% arrange(variable) 
viol_descriptions <- text3 %>% filter( variable %in% names(viol)) 
knitr::kable(viol_descriptions, caption = "Description of Variables in `viol`")
rm(text, text1, text2, text3)
```
